Applying Gaussian Process Regression to Predict Changes in Stock Price

### 2014-05-10 paper/paper.tex
Final paper

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage[T1]{fontenc}
%\usepackage{babel}

@@

\section{Introduction}
In stock markets, investors need long-term forecasting techniques to choose the right time to buy/sell stocks to maximize their profits or to minimize their loss. The majority of existing stock market forecasting techniques rely on machine learning tools and the accuracy of these tools is still a significant challenge for market traders considering the several external factors which affect the stock market. In this project we have tried to forecast change in price of a single stock using the Gaussian Processes (GP) model.

\section{Initial Data Processing}
Our data set (obtained from http://www.tworoads.co.in/trading-competition) consists of information collected on a single stock, including instantaneous stock prices and instantaneous values of 27 parameters ($P$) known to have some affect on the price.
These values are taken at many different time stamps.
Our goal is to predict the stock price in the near future, given current values of the parameters.

@@

Our algorithm could be divided into 4 major parts - preprocessing the data, building the covariance matrix, tuning the hyperparameters, learning the GP model using a windowing technique and finally predicting the change in stock price for the next time instance using the learned GP model.

@@

We experiment with 3 different strategies to build our training set from the available dataset. In the first strategy we prune out all data points that were more than one standard deviation away from the sample mean $\bf\bar{y}$. In the second strategy we prune out all data points that were within one standard deviation from the sample mean $\bf\bar{y}$. And for the third strategy, we keep all data points. After applying any one of the above strategies to the data, we linearly scale the inputs to have zero mean and unit variance on the training set. We also centered the corresponding target outputs so as to have zero mean on the training set.

@@

Then, the covariance functions are:
\begin{align}
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \exp{- \frac{1}{2} u^2} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Squared Exponential)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \exp{- u} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Ornstein–Uhlenbeck)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \left( 1 + \sqrt{3} u \right) \exp{- \sqrt{3}  u} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Matern - $\nu = \frac{3}{2}$)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \left( 1 + \sqrt{5} u \right) + 5 u^2) \exp{- \sqrt{5} u} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Matern  - $\nu = \frac{5}{2}$)}
\end{align}

@@

\subsection{Tuning the hyperparameters}
For the covariance matrices above, we tuned the hyperparameters by maximizing the log marginal likelihood, 
\begin{equation}
	\log p(\bfy | X, \bftheta) = - \frac{1}{2} \bfy^\top K_y^{-1} \bfy - \frac{1}{2} \log \| K_y \| - \frac{n}{2} \log (2 \pi)
\end{equation}

@@

We observed that the resultant $\bfl$ parameters had very large values in the majority of their coefficients and smaller values for a limited subset of components.
This means that a limited number of components have a larger correlation with the target value.

@@

We have applied a windowing technique to learn a GP on the data. We use the covariance matrix built using data from $t = 0$ to $t = \mathcal{T}$, to predict the change in stock price at time $t = \mathcal{T}+1$. Once the prediction is made, we use the input vector at $t = \mathcal{T}+1$ and the \textbf{\textit{predicted}} change in price at that instance as a new data point and compute the covariance matrix again for the window from $t = 1$ to $t = \mathcal{T}+1$. We use this newly computed covariance matrix to predict the change in stock price at $t=\mathcal{T}+2$. This process is repeated until $t = \mathcal{T}+T$, where $T$ denotes how far ahead in future we want to keep predicting the change in stock prices. Since re-computing the covariance matrix for every window can be time consuming, we compute only the last column / row of the covariance matrix as it corresponds to the time instance $t = \mathcal{T}+1$. In addition, we drop the row and column that correspond to the oldest time instance of the previous window, from the covariance matrix. This keeps the size of our covariance matrix constant (therefore, keeping the amount of computation required for inverting the covariance matrix constant across every window) while allowing us to re-use the other columns of the covariance matrix that correspond to the time instances common to the previous window and the current window.

@@

\subsection{Covariance Type}
For this experiment, we use $\mathcal{T} = 20000$, with Strategy 3, and try to predict 500 ms into the future.
We found that squared exponential and the Matern function with $\nu = \frac{5}{2}$ predicted nearly a straight line, with very small variations.
The Matern functio wiht $\nu = \frac{3}{2}$ and Ornstein–Uhlenbeck functions' predictions are shaped more like the actual data, with sharp peaks and valleys.

The Ornstein-Uhlenbeck model seems to fit the data the best, so we use it for all future experiments.

% Zoom plot in - 50-100??

\subsection{Window size}
Next, we experimented with varying window size $\mathcal{T}$, while using Strategy 3 and predicting 500 ms into the future.
We found that the larger windows predicted the dominant trend much better, and had fewer peaks and valleys.
However, the peaks and valleys that the larger windows predicted were, on average, more likely to be correct.

\subsection{Strategy Type}
We tried several different strategies (described in \ref{sec:strategy}).
\begin{table}[h]
	\label{t:miss}
	\centering
	\caption{Earnings from different strategies}
	\begin{tabular}{| l | l |}
	\hline
	Strategy & Earnings \\
	\hline
	1 & 42.88 \\
	2 & 28.39 \\
	3 & 3.86 \\
	\hline
	\end{tabular}
\end{table}

@@

If we had more time, we would run additional experiments further into the future, to see how far ahead we can predict before our predictions get notably worse.

\subsection{Comparisons to other models}
The dataset that we used came with linear regression and PCA regression models, trained on a much larger version of the same dataset.
Although the experiments are not trained and tested in the same way, we give a broad comparison to their results, to show that we performed similarly well.

\section{Conclusion}
The results from our experiments indicate that GP models could learn the correlations hidden among financial data to a very high degree when compared to other traditional ML algorithms like linear regression and PCA regression. But the biggest drawback of GP models as implemented in our application is the computation time that is required for the GP modeling. Regression based on GP models involve a huge amount of matrix inversions, the complexity of which is to the order of $\mathcal{O}(n^3)$. These computational needs, restricted the amount of training data that could be used, to a maximum of only 20000 training samples, even though we had access to around 2GB of stock price data that we would have liked to use during training. Several authors have suggested sparse approximation approaches for calculating the covariance matrix, which we believe is a promising direction for future work.

@@

\subsubsection*{References}
\small{
C.E. Rasmussen \& C.K.I. Williams, Gaussian Processes for Machine Learning. \\
David Mackay, Introduction to Gaussian Processes.\\
Long-term Stock Market Forecasting using Gaussian Processes, Nando de Freitas.
Dataset - http://www.tworoads.co.in/trading-competition}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
%\usepackage{babel}

@@

\section{Introduction}
In stock markets, investors need long-term forecasting techniques to choose the right time to buy/sell stocks to maximize their profits or to minimize their loss. The majority of existing stock market forecasting techniques rely on machine learning tools and the accuracy of these tools is still a significant challenge for market traders considering the several external factors which affect the stock market. In this project we have tried to forecast change in stock price using the Gaussian Processes (GP) model.

\section{Initial Data Processing}
Our data set (obtained from [4]) consists of information collected on a single stock, including instantaneous stock prices and instantaneous values of 27 parameters ($P$) known to have some effect on the price.
These values are taken at many different time stamps.
Our goal is to predict the stock price in the near future, given current values of the parameters.

@@

Our algorithm could be divided into 5 major parts - preprocessing the data, building the covariance matrix, tuning the hyperparameters, learning the GP model using a windowing technique and finally predicting the change in stock price for the next time instance using the learned GP model.

@@

We experiment with 3 different strategies to build our training set from the available dataset. In the first strategy we prune out all data points that were more than one standard deviation \textbf{away} from the sample mean $\bf\bar{y}$. In the second strategy we prune out all data points that were \textbf{within} one standard deviation from the sample mean $\bf\bar{y}$. And for the third strategy, we keep all data points. After applying any one of the above strategies to the data, we linearly scale the inputs to have zero mean and unit variance on the training set. We also centered the corresponding target outputs so as to have zero mean on the training set.

@@

Then, the covariance functions are:
\begin{align}
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \exp\{- \frac{1}{2} u^2\} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Squared Exponential)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \exp\{- u\} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Ornstein Uhlenbeck)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \left( 1 + \sqrt{3} u \right) \exp\{- \sqrt{3}  u\} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Matern - $\nu = \frac{3}{2}$)}   \\
	k(\bfx_p, \bfx_q) &= \sigma_f^2 \left( 1 + \sqrt{5} u \right) + 5 u^2) \exp\{- \sqrt{5} u\} + \sigma_n \delta_{pq} \quad \quad \quad \mbox{(Matern  - $\nu = \frac{5}{2}$)}
\end{align}

@@

\subsection{Tuning the hyperparameters}
For the covariance matrices above, we tuned the hyperparameters by maximizing the log marginal likelihood, 
\begin{equation}
	\log p(\bfy | X, \bftheta) = - \frac{1}{2} \bfy^\top K_y^{-1} \bfy - \frac{1}{2} \log | K_y | - \frac{n}{2} \log (2 \pi)
\end{equation}

@@

We also observed that the resultant $\bfl$ parameters had very large values in the majority of their coefficients and smaller values for a limited subset of components.
This means that a limited number of components have a larger correlation with the target value.

@@

We have applied a windowing technique to learn a GP on the data. We use the covariance matrix built using data from $t = 0$ to $t = \mathcal{T}$, to predict the change in stock price at time $t = \mathcal{T}+1$. Once the prediction is made, we use the input vector at $t = \mathcal{T}+1$ and the \textbf{\textit{predicted}} change in price at that instance as a new data point and compute the covariance matrix again for the window from $t = 1$ to $t = \mathcal{T}+1$. We use this newly computed covariance matrix to predict the change in stock price at $t=\mathcal{T}+2$. This process is repeated until $t = \mathcal{T}+T$, where $T$ denotes how far ahead in future we want to keep predicting the change in stock prices.

Since re-computing the covariance matrix for every window can be time consuming, we compute only the last column / row of the covariance matrix as it corresponds to the time instance $t = \mathcal{T}+1$. In addition, we drop the row and column that correspond to the oldest time instance of the previous window, from the covariance matrix. We repeat this procedure for every new $t$ that we predict. This keeps the size of our covariance matrix constant (therefore, keeping the amount of computation required for inverting the covariance matrix constant across every window) while allowing us to re-use the other columns of the covariance matrix that correspond to the time instances common to the previous window and the current window.

@@

\subsection{Covariance Type}
For this experiment, we use $\mathcal{T} = 20000$, with Strategy 3, and try to predict 500 ms into the future.
In figure \ref{fig1}, we found that squared exponential, Ornstein Uhlenbeck and the Matern function with $\nu = \frac{5}{2}$ predicted nearly a straight line, with very small variations.
The Matern function with $\nu = \frac{3}{2}$ had predictions shaped more like the actual data, with sharp peaks and valleys.

\begin{figure}[h]
	\label{fig1}
	\centering
	\includegraphics[width=0.8\textwidth]{../code/fig_1.png}
	\caption{Covariance Functions Experiment}
\end{figure}

The Matern function with $\nu = \frac{3}{2}$ looked most like the data, but the Ornstein-Uhlenbeck model had a smaller mean squared error, so we used it for all future experiments.

\subsection{Window size}
Next, we experimented with varying window size $\mathcal{T}$, while using Strategy 3 and predicting 500 ms into the future.
In figure \ref{fig2}, we found that the larger windows predicted the dominant trend much better, and had fewer peaks and valleys.

\begin{figure}[h]
	\label{fig2}
	\centering
	\includegraphics[width=0.8\textwidth]{../code/fig_2.png}
	\caption{Window Experiments}
\end{figure}

\subsection{Strategy Type}
We tried several different strategies (described in \ref{sec:strategy}).
We then applied an earnings predictor, provided with [4], to our predicted values.
The earnings of a model measure how well it fits the dataset.
\begin{table}[h]
	\label{t}
	\centering
	\caption{Earnings from different strategies}
	\begin{tabular}{| l | l |}
	\hline
	Strategy & Earnings \\
	\hline
	1 & 42.88 \\
	2 & 28.39 \\
	3 & 3.86 \\
	\hline
	\end{tabular}
\end{table}

In table \ref{t}, we found that breaking the data into inliers and outliers helped us to predict the trend much better.
We think this is because these models had less variation and were easier to train.
With more training data and time, the Strategy 0 model may have done better too.

@@

If we had more time, we would run additional experiments further into the future, to see how far ahead we can predict before our model's performance gets notably worse.
We also discovered that the data does not have a high enough precision.  In many cases, the stock prices aren't given to enough decimal places, so the change in price is just 0.
Using larger time gaps might have helped to ameliorate this problem.

\subsection{Comparisons to other models}
The dataset that we used came with pre-trained linear regression and PCA regression models.
In Figure \ref{fig4}, we show comparisons of earnings of the different learning algorithms.  Each group of bars is tested on the same set of data.

\begin{figure}[h]
	\label{fig4}
	\centering
	\includegraphics[width=0.8\textwidth]{../code/fig_4.png}
	\caption{Comparisons Between Different Learning Algorithms}
\end{figure}

\section{Conclusion}
The results from our experiments indicate that GP models could learn the correlations hidden among financial data to a very high degree when compared to other traditional ML algorithms like linear regression and PCA regression. But the biggest drawback of GP models as implemented in our application is the computation time that is required for the GP modeling. Regression based on GP models involve a huge amount of matrix inversions, the complexity of which is to the order of $\mathcal{O}(n^3)$. These computational needs, restricted the amount of training data that could be used, to a maximum of only 20000 training samples, even though we had access to around 2GB of stock price data that we would have liked to use during training. Several authors have suggested sparse approximation approaches for calculating the covariance matrix, which we believe is a promising direction for future work.
Another improvement would be to select the predicted values by sampling from the distribution, rather than using the distribution mean.

\subsubsection*{Contributions}
Saswati contributed the experiment design and the Gaussian Processes code.
Stephanie contributed the covariance optimization code.
Both team members ran experiments, contributed to the paper, and researched the models and methods to be used.

@@

\subsubsection*{References}
\small{
	[1] C.E. Rasmussen \& C.K.I. Williams, Gaussian Processes for Machine Learning.

	[2] David Mackay, Introduction to Gaussian Processes.

 [3] Long-term Stock Market Forecasting using Gaussian Processes, Nando de Freitas.

 [4] Dataset - http://www.tworoads.co.in/trading-competition
 }
